# Workshop 10: Logistic regression: Diagnostics etc.

:::: {.blackbox data-latex=""}
::: {.center data-latex=""}
**Learning outcomes**
:::

<ul>
  <li>Adding multiple predictors to a logistic regression in *R*</li>
  <li>Prediction in logistic regression</li>
  <li>Coding Average Predictive Comparisons</li>
</ul>
::::

## Multiple predictors

Today we'll be using the data set "dmd.csv" and run a regression with multiple predictors. There are 209 observations corresponding to blood samples collected in a project to develop a screening program for female relatives of boys with Duchenne Muscular Dystrophy (DMD) a congenital disease. There are 4 questions we need to answer:

1. The program's main goal was to inform a woman of her chances of being a carrier based on serum markers (chemicals in the blood). 
2. Also of interest was whether age should be taken into account. 

Enzyme levels were measured in known carriers (75 samples) and in a group of non-carriers (134 samples). The first two serum markers, creatine kinase and hemopexin (``ck``,``h``), are inexpensive to measure, while the last, pyruvate kinase (``pk``) is more expensive to measure. 

3. We also want to know whether ``ck`` and ``h`` are sufficient to predict carrier status or whether ``pk`` is necessary. In routine blood tests, it is not possible to measure all three, if we measure ``ck`` and ``h`` we cannot measure ``pk``. However all three were measured for this study.

The water supply in the lab in which the blood samples were being analysed changed during the study. ``afterchg=1`` indicates which samples were taken after the change and  ``afterchg=0`` which were taken prior to the change.

4. Finally we want to know whether the change in water had an effect on the outcomes

The data look like this:

```{r}
dmd.dat<-read.csv("../Data/dmd2.csv",header = T)
head(dmd.dat)
```

Before we do any regressions let's centre the continuous predictors. This makes sense because the measurements are not taken on new-borns and 0 values are not possible for the serum markers as they exist in some concentration in every person. 

```{r}
#write function
cent.pred<-function(v){
  cent.v<-v-mean(v)
  cent.v
}  
#center
cent.ck<-cent.pred(dmd.dat$ck)
cent.h<-cent.pred(dmd.dat$h)
cent.age<-cent.pred(dmd.dat$age)
cent.pk<-cent.pred(dmd.dat$pk)
```

Let's start by using two regressions, one with ``cent.pk``,

```{r}
dmd.pk.glm<-glm(carrier~cent.age+afterchg+cent.pk,data=dmd.dat,
                family=binomial(link="logit"))
display(dmd.pk.glm)
```

and another with ``cent.ck``, ``cent.h``.

```{r}
dmd.ckh.glm<-glm(carrier~cent.age+afterchg+cent.ck+cent.h,data=dmd.dat,
                 family=binomial(link="logit"))
display(dmd.ckh.glm)
```

``pk``, ``ck`` and ``h`` are all significant, indicating that the models _do_ help predict carrier status. We can confirm this using the diagnostics: For both models the difference between the null and residual deviances are above the critical value for the corresponding $\chi^2$ distribution. Specifically:

1. For the regression with ``pk`` the difference between the residual and null deviance is 109.9. This needs to be compared to the $\chi^2$ distribution on 3 degrees of freedom. 
```{r}
qchisq(0.95,3)
```
109.9>7.81 so the model with ``pk``, ``age`` and ``afterchg`` is better than the null model. A similar argument leads us to compare 120.4 (difference between null and residual deviance in model with ``ck`` and ``h``) with:
```{r}
qchisq(0.95,4)
```
So the model with  ``ck``,  ``h``, ``age`` and ``afterchg`` is better than the null model.

However in both cases ``afterchg`` appears to have little effect in terms of significance Furthermore, ``age`` is a significant predictor of being a carrier. This answers questions 1,2 and 4 above.

Let's turn to question 3: "Are ``ck`` and ``h`` sufficient to predict carrier status, or is ``pk`` better?" In order to do this, let's compare the classification tables for both models. First let's write a function that produces classification tables.

```{r tidy=TRUE}
#function
ct.op<-function(predicted,observed){ #arguments
#create the data frame  
df.op<-data.frame(predicted=predicted,observed=observed)
#create a table 
op.tab<-table(df.op)
#use the prop.table function to obtain the proportions we need:
#those who were correctly predicted as 0 
#@position 1,1 in the table of proportions
obs0.tab<-round(prop.table(op.tab,2)[1,1],2)
#those who were correctly predicted as 1
#@position 2,2 in the table of proportions
obs1.tab<-round(prop.table(op.tab,2)[2,2],2)
#and put them under the table 
op.tab<-rbind(op.tab,c(obs0.tab,obs1.tab))
#name the rows
rownames(op.tab)<-c("pred=0","pred=1","%corr")
#name the columns
colnames(op.tab)<-c("obs=0","obs=1")
#return the table
op.tab
}
```

How well does the model with ``pk`` predict?

```{r}
#we use as.numeric because the part inside the brackets returns TRUE/FALSE
pred.pk<-as.numeric(dmd.pk.glm$fitted.values>0.5)
#pass the fitted values and the observed values to ct.op
ct.op(pred.pk,dmd.dat$carrier)
```

How about the model with ``ck`` and ``h``?

```{r}
pred.ckh<-as.numeric(dmd.ckh.glm$fitted.values>0.5)
ct.op(pred.ckh,dmd.dat$carrier)
```


:::: {.whitebox data-latex=""}
Q: Based on the output of *display()* and the tables of predicted vs observed would you recommend the test using *ck* and *h* or the test using *pk*? Explain your answer.
::::

## Predicting probabilities and risk ratios

As mentioned before, the logistic function is non-linear which means that a change in one in the value of a predictor is not the same across the range of the predictor. Specifically, ``age`` going from 20-30 may not result in the same change in the probability of being a carrier as ``age`` going from 30-40.

:::: {.whitebox data-latex=""}
Q:Using plots, explain the statement above.
::::

One way to see this numerically is to look at _risk ratios_  or _relative risks_. We've done this by hand for a single predictor, but how can we get ``R`` to do it for us in the context of multiple predictors?

Consider the model with ``ck``, ``h`` and ``age``:

```{r}
dmd.glm.apc<-glm(carrier~age+ck+h,data=dmd.dat,family=binomial)
display(dmd.glm.apc)
```

What is the change in probabilities associated with a change in _age_ of 

a) 20 to 25 and 
b) 40 to 45 

for a woman with mean ``ck`` (90.19) and mean ``h`` (86.37)? Ages over 45 are not of interest because the probability of women above 45 having children is very low.

1. We create the data frame we want to use for prediction: Note that you could create this manually in excel and save it as a csv file and load it into ``R`` if you prefer.

```{r}
new.dmd.dat<-data.frame(age=c(20,25,40,45),ck=mean(dmd.dat$ck),h=mean(dmd.dat$h))
new.dmd.dat

```

2. Next we use the predict function, this gives us the logits associated with the values above.

```{r}
pred.logit<-predict(dmd.glm.apc,new.dmd.dat)
#pred.logit
```
 3. To obtain probabilities, we can either use the formula or we can ask ``R`` to do it: (remember the ``invlogit`` function is in the ``arm`` package.)
 
```{r}
pred.probs<-invlogit(pred.logit)
#pred.probs
```
 
To see these together a small data frame:

```{r}
logreg<-data.frame(logits=pred.logit,probs=pred.probs)
cbind(new.dmd.dat,logreg)
```

 We see that the probabilities of being a carrier are very low for younger women and increase substantially for older women with mean values of ``ck`` and ``h``.
 
 4. Now calculate the risk ratios: 
 
```{r}
print("risk ratio age:20-25, mean ck and h")
as.numeric(pred.probs[2]/pred.probs[1])
print("risk ratio age:40-45, mean ck and h")
as.numeric(pred.probs[4]/pred.probs[3])
#as.numeric means we don't get the data frame formatting
```

We see that the probability of being a carrier increases by 80% when going from the age of 20 to 25 whereas is only increases by 22% when going from 40-45 years of age for women with average ``ck`` and average ``h`` levels. These figures look alarming without the probabilities themselves to relate to so display risk ratios alongside probabilities.
 
:::: {.whitebox data-latex=""}
Q: Why do is it a good idea to use the mean (or median) value for the predictors we are not interested in comparing?
::::

 Let's see how those values are affected if we look at women with high ``ck``: 73 is the 75th quantile.

```{r}
new.dmd.dat<-data.frame(age=c(20,25,40,45),ck=73,h=mean(dmd.dat$h))
pred.logit<-predict(dmd.glm.apc,new.dmd.dat)
pred.probs<-invlogit(pred.logit)
logreg<-data.frame(logits=pred.logit,probs=pred.probs)
cbind(new.dmd.dat,logreg)
print("risk ratio age:20-25, ck=73 and  mean h")
as.numeric(pred.probs[2]/pred.probs[1])
print("risk ratio age:40-25, ck=73 and  mean h")
as.numeric(pred.probs[4]/pred.probs[3])
```

For women with high``ck`` the situation appears to be only slightly worse in terms of risk ratios (85 and 29% for younger and older women respectively) but is actually slightly better in terms of the probabilities (13 and 71% probability of being a carrier at 25 and 45 years respectively). 

## Coding APCs

The APC is a generalisation of the approach above where we input predictor means for the predictors that are not the focus of our interest. One big difference is that we look at _differences_ between probabilities and _not ratios_.

Let's apply this technique to the DMD data set. We should not use centered predictors for this. Also, remove ``afterchg`` as it is not significant.

For age:
```{r}
dmd.glm.apc<-glm(carrier~age+ck+h,data=dmd.dat,family=binomial)

b<-coef(dmd.glm.apc)
hi<-45 #actual highest is 61 but women mostly have children before 45
lo<-20
delta.age<-with(dmd.dat,(invlogit(b[1]+b[2]*hi+b[3]*ck+b[4]*h) 
                         -invlogit(b[1]+b[2]*lo+b[3]*ck+b[4]*h)))
print(mean(delta.age))              
```

This means that the average predictive difference in probability of being a carrier associated with an increase over the observed range of years is 46% other things remaining equal. Remember that the range of age is from 20-61 but we only consider 20-45 as child bearing ages.

For ``ck``:
```{r}
hi<-1000
lo<-15
delta.ck<-with(dmd.dat,(invlogit(b[1]+b[2]*age+b[3]*hi+b[4]*h)
                        -invlogit(b[1]+b[2]*age+b[3]*lo+b[4]*h)))
mean(delta.ck)              
```

For ``h``:

```{r}
hi<-118
lo<-34
delta.h<-with(dmd.dat,(invlogit(b[1]+b[2]*age+b[3]*ck+b[4]*hi)
                       -invlogit(b[1]+b[2]*age+b[3]*ck+b[4]*lo)))
print(mean(delta.h))              
```

:::: {.whitebox data-latex=""}
Q: Interpret the APC for *ck* and *h*.
::::



