# Appendix {-}

## Appendix A: ``ggplot2`` package 

:::: {.blackbox data-latex=""}
::: {.center data-latex=""}
**Learning outcomes**
:::

In this section there are example of plots using the ``ggplot2`` package. 
<ul>
  <li>Basic box and scatterplots</li>
  <li>Arranging plots using ``gridExtra``</li>
  <li>Adding points to a scatterplot</li>
  <li>Using colour, shape and fill to add information about categorical variables to a box and scatterplot</li>
  <li>Adding a straight line to a scatterplot</li>
  <li>Adding a linear model to a scatterplot in simple linear regression</li>
  <li>Adding multiple linear models to a scatterplot in multiple linear regression, without an interaction and with ith an interaction</li>
  <li>Using ``facet_wrap`` to display data across categories</li>
  <li>Barplots for cross tabulation</li>
  <li>Special plots</li>
</ul>
::::

Throughout we'll use the same data: The Boston Housing data. This has been modified so that ``ZN``, one of the predictors is a 3 level categorical predictor.

```{r}
Boston.dat<-read.csv("../Data/Bostonggplot.csv", header=TRUE)
head(Boston.dat,3)
```

You will need the following packages:

```{r}
library(ggplot2)
library(gridExtra)
```


### Basic box and scatterplots

Let's consider the continuous predictor ``LSTAT``.

```{r fig.show='hide'}
#aes stands for "aesthetics"
#without +geom_point() the plot is empty
ggplot(data=Boston.dat, aes(x=LSTAT, y=MV))+geom_point()
```

```{r fig.show='hide'}
#aes stands for "aesthetics"
#without +geom_boxplot() the plot is empty
ggplot(data=Boston.dat, aes(x=ZN, y=MV))+geom_boxplot()
```

### Arranging plots to save space 

To avoid using a lot of space, use ``grid.arrange()`` in the ``gridExtra`` package.

```{r fig.show='hide'}
#scatterplot
p1<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV))+geom_point()
#boxplot
p2<-ggplot(data=Boston.dat, aes(x=ZN, y=MV))+geom_boxplot()
#place them side by side (nrow=1) or one above the other (nrow=2)
grid.arrange(p1,p2,nrow=1)
```

Most changes you want to make to the scatter/boxplot are now just a matter of adding something to ``p1``/``p2`` . Some changes have to be in the main body of ``ggplot()`` in aes``()``.

```{r fig.asp=0.4}
#scatterplot with different colours for different levels of ZN
p1<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV, colour=ZN))+geom_point()
#boxplot with colour in the boxplots
p2<-ggplot(data=Boston.dat, aes(x=ZN, y=MV,fill=ZN))+geom_boxplot()
#place them side by side (nrow=1) or one above the other (nrow=2)
grid.arrange(p1,p2,nrow=1)
```

### Adding points and lines

Let's add

1. a point and a straight line to the scatterplot
2. An interaction to a boxplot

```{r}
#basic scatterplot
p1<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV, colour=ZN))+geom_point()
#creating point
new.point<-data.frame(LSTAT=30,MV=40,ZN="Zero")
#adding point in red with a larger size and a different shape
p1<-p1+geom_point(data=new.point, color="red", size=3, shape=3)
#adding a straight line with intercept 50 and slope -2
p1<-p1+geom_abline(intercept=50, slope=-2)

#basic boxplot with ZN grouped by CHAS
p2<-ggplot(data=Boston.dat, aes(x=ZN, y=MV,fill=factor(CHAS)))+geom_boxplot()
#over two rows for grid.arrange
grid.arrange(p1, p2, nrow=2)
```

Next, let's add
    
    1. A line of best fit to a scatterplot
    2. Add points to the boxplot

```{r fig.asp=0.4}
#basic scatterplot
p1<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV))+geom_point()
#adding a line of best fit using a linear model (se=FALSE means no error bars)
p1<-p1+geom_smooth(method="lm",se=FALSE)

#basic boxplot with ZN 
p2<-ggplot(data=Boston.dat, aes(x=ZN, y=MV))+geom_boxplot()
#add the points and jitter them so they group together too much
p2<-p2+geom_jitter(shape=16, position=position_jitter(0.2))

grid.arrange(p1, p2, nrow=1)
```

### Adding a regression line

If we add a ``geom_smooth`` to the plot with ``colour=ZN``, three non-parallell lines appear. Basically ``geom_smooth`` uses the interaction model to plot the lines. If we want to show what the lines look like without an interaction then we need to use ``geom_abline()``


```{r fig.asp=0.4, tidy=TRUE}
#basic scatterplot
p1<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV, color=ZN))+geom_point()
#adding a line of best fit using a linear model (se=FALSE means no error bars)
p1<-p1+geom_smooth(method="lm",se=FALSE)

#run the regression:
lm.plot<-lm(MV~LSTAT+ZN, data=Boston.dat)
#basic  scatterplot
p2<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV, colour=ZN))+geom_point()
#add three lines 
p2<- p2+ geom_abline(intercept = coef(lm.plot)[1], slope = coef(lm.plot)[2], color="red")
p2<- p2 + geom_abline(intercept = (coef(lm.plot)[1]+ coef(lm.plot)[3]), 
                      slope = coef(lm.plot)[2],color="green")
p2<- p2 + geom_abline(intercept = (coef(lm.plot)[1]+ coef(lm.plot)[4]), 
                      slope = coef(lm.plot)[2],color="lightblue") 

grid.arrange(p1, p2, nrow=1)
```

### Using ``facet_wrap()``

Another way of displaying data by levels of a categorical variable is to use ``facet_wrap()``.

```{r fig.asp=0.4}
#basic scatterplot
p1<-ggplot(data=Boston.dat, aes(x=LSTAT, y=MV, color=ZN))+geom_point()
#adding a line of best fit using a linear model (se=FALSE means no error bars)
p1<-p1+geom_smooth(method="lm",se=FALSE)
p1<-p1+facet_wrap(~ZN)
p1
```

### Barplots

The barplot for the crosstabulation of two categorical variables looks like this:

```{r}
p1<- ggplot(data=Boston.dat,aes(x = factor(CHAS), fill = ZN)) +
  geom_bar(position = "fill") +  scale_y_continuous(name = "Within group Percentage"
                     , labels = scales::percent)
p1

```

### Special plots

Below are a few plots that require special code that we don't cover in depth in the course.

```{r warning=FALSE}
special.dat<- gather(data=Boston.dat[,-c(4,2)], -MV, key="var", value="value") 
#gather (and similar functions like melt) stack 
#the continuous predictor variables on top of each other. 
#-MV means that MV stays the same
#This allows us to use facet_wrap when the variables have different attributes
#We remove columns 2 and 4 as these are the categorical variables 
#To see the data type:
#View(special.dat)
#now for the plot using special dat and scales="free_x"
#free_x means that the x-axis is not the same for each one of the plots
p1<-ggplot(special.dat, aes(x = value, y = MV, colour=var))+geom_point(size=0.5)+geom_smooth(method="lm", se=FALSE, color="black") +facet_wrap(~ var, scales = "free_x")
#legend is not useful in these multi-variable plots
p1<-p1+theme(legend.position = "none")
p1

#categorical predictors (two in this case)
special.dat<- gather(data=Boston.dat[,c(2,4,13)], -MV, key="var", value="value") 
# fill=factor(value) means that each boxplot will have a different colour
p2<-ggplot(special.dat, aes(x = factor(value), y = MV,fill=factor(value)))+geom_boxplot() +facet_wrap(~ var, scales = "free_x")
p2<-p2+theme(legend.position = "none")
p2
```

